---
title: "Gov 50 Final Project"
author: "Olivia Ma"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r}
library(tidyverse)
cj <- read_csv("data/response.csv")
```

## 1. Introduction

Criminal justice and related policies always have been the center of U.S. political debates and elections. In the most recent 2022 U.S. midterm elections, topics such as legalization of marijuana, criminalization of abortion, and prison labor have all caught public attention (https://www.themarshallproject.org/2022/11/12/7-key-criminal-justice-takeaways-from-the-midterms). Yet very few policy makers have personal experience with incarceration. Then, for the incarcerated, how do stances on topics related to criminal justice relate to a change in the criminals' political ideologies after incarceration? In this report, I examine this question by analyzing the correlation between criminals' stances on political issues and whether their political ideologies have changed after incarceration. For further elaboration, I specifically examines the criminals' stances on assault weapon ban, marijuana legalization, tightening border security, and raising minimum wages and how they relate to whether their political ideologies have changed after incarceration. My null hypothesis is that political stances on the aforementioned issues will not contribute to the change in political ideologies after incarceration. Through this report, I hope to bring in a new perspective on criminal justice, one coming from the point of views of the incarcerated. 

## 2. Data Description 

This project is based on the Marshall Project's ongoing survey of the political views of incarcertated people in American prisons and jails: 

> Bartley, Lawrence, Nicole Lewis, and Anna Flagg. 2017. "[How We Pulled Off A Groundbreaking Political Survey Behind Bars.](https://www.themarshallproject.org/2020/03/11/how-we-pulled-off-a-groundbreaking-political-survey-behind-bars)" The Marshall Project.

This is an observational study generated through a survey sent out to some prisons and jails in the U.S. in 2019. Responses were collected on a rolling basis until March, 2020. This survey received a total of 8266 responses. However, the survey was not able to reach all demographic and geographic regions, due to the limitation of collecting data from incarcerated population. 35 questions were asked in this survey, ranging from demographic information to issues related to criminal justice. In this report, I selected the following variables to examine the correlation in between.

The data file for this study is `data/response.csv` and contains the following variables:

| Name                      | Description                                               |
|:--------------------------|:----------------------------------------------------------|
| `party`                               | the respondent's party affiliation                                        |
| `stance_on_assault_weapons_ban`       | the respondent's opinion on assault weapons ban: "Strongly support", "Support", "Undecided", "Oppose", "Strongly oppose" |
| `stance_on_marijuana_legalization`    | the respondent's opinion on marijuana legalization: "Strongly support", "Support", "Undecided", "Oppose", "Strongly oppose" |
| `stance_on_tightening_border_security`| the respondent's opinion on tightening border security: "Strongly support", "Support", "Undecided", "Oppose", "Strongly oppose" |
| `stance_on_raise_min_wage`            | the respondent's opinion on raising minimum wage: "Strongly support", "Support", "Undecided", "Oppose", "Strongly oppose" |
| `politics_changed_since_incarcerated` | whether the respondent's political ideology has changed since incarceration; Yes", "No", "NA" if withheld |

### Tidying data
In the following section, I selected the aforementioned variables and tidied the data. I simplified the names of these variables, and numericalized the categorical variables. "Strongly support", "Support", "Undecided", "Oppose", "Strongly oppose" are represented by numeric values `2`, `1`, `0`, `-1`, `-2`. "Yes", "No", "NA" are I categorized into `1` for "Yes" and `0` for "No", with "NA" dropped. 

```{r}
cj_tidy <- cj |>
  mutate(weapons_ban = case_when(
           (stance_on_assault_weapons_ban == "Undecided") ~ 0, 
           (stance_on_assault_weapons_ban == "Oppose") ~ -1, 
           (stance_on_assault_weapons_ban == "Strongly oppose") ~ -2,
           (stance_on_assault_weapons_ban == "Support") ~ 1,
           (stance_on_assault_weapons_ban == "Strongly support") ~ 2,
         ),
         marijuana = case_when(
           (stance_on_marijuana_legalization == "Undecided") ~ 0, 
           (stance_on_marijuana_legalization == "Oppose") ~ -1, 
           (stance_on_marijuana_legalization == "Strongly oppose") ~ -2,
           (stance_on_marijuana_legalization == "Support") ~ 1,
           (stance_on_marijuana_legalization == "Strongly support") ~ 2,
         ),
         border_security = case_when(
           (stance_on_tightening_border_security == "Undecided") ~ 0, 
           (stance_on_tightening_border_security == "Oppose") ~ -1, 
           (stance_on_tightening_border_security == "Strongly oppose") ~ -2,
           (stance_on_tightening_border_security == "Support") ~ 1,
           (stance_on_tightening_border_security == "Strongly support") ~ 2,
         ),
        min_wage = case_when(
           (stance_on_raise_min_wage == "Undecided") ~ 0, 
           (stance_on_raise_min_wage == "Oppose") ~ -1, 
           (stance_on_raise_min_wage == "Strongly oppose") ~ -2,
           (stance_on_raise_min_wage == "Support") ~ 1,
           (stance_on_raise_min_wage == "Strongly support") ~ 2,
         ),
         change = if_else(politics_changed_since_incarcerated == "Yes", 1, 0)) |>
  select(party, change, weapons_ban, marijuana, border_security, min_wage) |>
  drop_na()
cj_tidy

change_bar <- cj_tidy |>
  ggplot(aes(x = factor(change))) + geom_bar() +
  labs(title = "Bar Plot Distribution of Reponsdents Political Ideology Changes",
       x = "Change in Political Ideology",
       y = "Count of Change in Political Ideologies") +
  scale_x_discrete(labels = c("No", "Yes"))
change_bar
```

## 3. Result: Regressions
### Model 1
This regression examines what beliefs on criminal justice may contribute to the change in political ideologies after incarceration.
Null hypothesis: Political beliefs on weapon ban, marijuana legalization, tightening border security, and raising minimum wage will not contribute to the change in political ideologies after incarceration. 

After conducting a linear regression with `change` as my outcome variable, and `weapons_ban`, `marijuana`, `border_security`, and`min_wage`as my explanatory variable, I can either reject or not reject my null hypothesis (whether political ideologies changed or not) based on if the p-value is small enough. I can observe whether political stances will positively or negatively impact `change`. If the linear regression has a positive slope, it means that political stances will positively impact `change`, and vice versa. 

```{r}
library(broom)
library(knitr)

change_multi <- lm(change ~ weapons_ban + marijuana + border_security + min_wage, data = cj_tidy)
change_multi |>
  tidy() |>
  knitr::kable(caption = "Coefficient Estimates for a Model Fitted to Estimate Variation in Change of Political Ideology",
               col.names = c("Predictor", "Estimate", "Std. Error", "Statistic", "P Value"),
               digits = 3)
summary(change_multi)

tidy(change_multi, conf.int = TRUE, conf.level = 0.95)|>
  knitr::kable(digits = 3)
```
Since the response variable (change in political ideology) is a binary variable, the point estimate of 0.459 is the average outcome when all other predictors (stance on tighten weapon bans, marijuana legalization, tighten border security, and raise minimum wage) equal to 0. This suggest that even when a prisoner is undecided on all these four issues, there is on average a 46% of chance that his/her political ideology will change after incarceration. When the prisoner is undecided on marijuana legalization, tighten border security, and raise minimum wage, the more strongly he/she support assault weapons ban (the support increase by 1 unit), the predicted change in political ideologies increase by approximately 3.5%. Similar analysis can be applied to marijuana legalization (predicted change in political ideologies decreases by 0.2%), tighten border security (predicted change in political ideologies increases by 1.3%), and raise minimum wage (predicted change in political ideologies increases by 1.7%). Hence, opinions on assault weapon bans has the highest impact on the predicted change in political ideologies after incarceration. However, this regression model can be further improved, as it has a R-squared value of 0.014, suggesting that only 1.4% of the observed variations is accounted for in this regression model. 

After conducting a hypothesis testing on the 95% confidence interval with the alpha-value being 0.05, we can conclude the following evaluation of the p-value: 
When the prisoners' beliefs on marijuana legalization, tightening border security, and raising minimum wage remain the same, there is strong enough evidence to reject the null hypothesis and thus we can state that beliefs on weapon ban will contribute to the change in political ideology (p-value for `weapons_ban` is `0`)
When the prisoners' beliefs on weapon ban, tightening border security, and raising minimum wage remain the same, there is not strong enough evidence to reject the null hypothesis and thus we can state that beliefs on marijuana legalization ban will not contribute to the change in political ideology (p-value for `marijuana` is `0.773`)
When the prisoners' beliefs on weapon ban, marijuana legalization, and raising minimum wage remain the same, there is strong enough evidence to reject the null hypothesis and thus we can state that beliefs on tightening border security ban will contribute to the change in political ideology (p-value for `border_security` is `0.002`)
When the prisoners' beliefs on marijuana legalization, tightening border security, and weapon ban remain the same, there is strong enough evidence to reject the null hypothesis and thus we can state that beliefs on raising minimum wage will contribute to the change in political ideology (p-value for `min_wage` is `0.003`)

In conclusion, in this multiple regression model, there is strong enough evidence to support that the prisoners' political stances on weapon bans, tightening border security, and raising minimum wages will contribute to the change in political ideologies, whereas there is not strong enough evidence to state that their stances on marijuana legalization will. However, since this is an observational study and not a controlled randomized experiment, this conclusion cannot be generalized.

```{r}
marijuana_bar <- cj_tidy |>
  ggplot(aes(x = factor(marijuana))) + 
  geom_bar() +
  labs(title = "Bar Plot Distribution of Respondents' Support of Marijuana Legalization",
       x = "Support of Marijuana Legalization",
       y = "Count of Change in Political Ideologies") +
  scale_x_discrete(labels = c("Strongly Oppose", "Oppose", "Undecided", "Support", "Strongly Support"))
marijuana_bar
```
To further examine 

### Model 2: Accounting for Party Afflication 
The previous regression model only measures whether the criminals' political beliefs have changed after incarceration. Hence, this multiple regression model accounts for their current party affiliation and examines whether their current party affiliation contributes to their change in political ideologies.
Null hypothesis: Party affiliation is not related to the change in criminals' political ideologies after incarceration
`party` is a categorical value measured in "Other", "The Democratic party", "The Republican party", and "Independent". "Independent" is the baseline in this model. 

```{r}
party_multi <- lm(change ~ weapons_ban + marijuana + border_security + min_wage + factor(party), data = cj_tidy)
party_multi |>
  tidy()

summary(party_multi)

tidy(party_multi, conf.int = TRUE, conf.level = 0.95)|>
  knitr::kable(digits = 3)
```

When the prisoners' four political beliefs (`weapons_ban`, `marijuana`, `border_security`, and `min_wage`) equal to 0 (i.e. they are undecided about all these four issues), the predicted change of party affiliation percentage from Independent to Other is -8.36%, to Democratic party is -1.79%, and to Republican party is 1.71%. This means that criminals will be 8.36% less like to be in other parties than Independent, 1.79% less likely to be in the Democratic party, and 1.71% more likely to be in the Republican party after incarceration. 

After conducting the hypothesis testing on a 95% confidence interval, we can conclude the following based on the p-value:
When the prisoners' beliefs on weapon ban, marijuana legalization, tightening border security, and raising minimum wage remain the same, there is strong enough evidence to reject the null hypothesis and thus state being in the Other party is related to the change in criminals' political ideologies after incarceration (p-value for `factor(party)Other` is `8.88e-07`)
When the prisoners' beliefs on weapon ban, marijuana legalization, tightening border security, and raising minimum wage remain the same, there is not strong enough evidence to reject the null hypothesis and thus we can state being in the Democratic party is not related to the change in criminals' political ideologies after incarceration (p-value for `factor(party)The Democratic party` is `0.24164`)
When the prisoners' beliefs on weapon ban, marijuana legalization, tightening border security, and raising minimum wage remain the same, there is not strong enough evidence to reject the null hypothesis and thus we can state being in the Republican party is not related to the change in criminals' political ideologies after incarceration (p-value for `factor(party)The Republican party` is `0.28390`)

In conclusion, in this multiple regression model, there is strong enough evidence to support that being in the Other party now is related to the change in political ideologies compared to being in the Independent party, whereas there is not strong enough evidence to state that being in the Democratic and Republican parties is related to the change in political ideologies compared to being in the Independent party. 


## instruction 
A results section, including a presentation of your regression results (either a plot or a table), a separate plot of the main relationship of interest (e.g. a scatterplot of the outcome and the main independent variable), and as well as a summary and discussion of results in substantive context (a paragraph or two). 

You should also interpret (in plain English) the main coefficient of interest in your regression, comment on the statistical significance of the estimated coefficient, and  say whether or not you believe the coefficient to represent a causal effect.

A conclusion section, including a recap of results, discussion of limitations of the study, and how to improve the study in the future (a paragraph)


(3) a results section that contains a scatterplot, barplot, or boxplot of the main relationship of interest and output for the main regression of interest; and 
(4) a brief (one paragraph) concluding section that summarizes your results, assesses the extent to which you find support for your hypothesis, describes limitations of your analysis and threats to inference, and states how your analysis could be improved (e.g., improved data that would be useful to collect).

